Parallel Hierarchical Clustering using \\ Rank-Two Nonnegative Matrix Factorization \thanks{Identify applicable funding agency here. If none, delete this.}

### 2020-07-14 latex/paper.tex
added description of parallel algorithsm

The communication cost of the algorithm depends on the parallel distribution of the input matrix data $\M{A}$.
In order to avoid redistribution of the matrix data, we choose a 1D row distribution so that each processor owns a subset of the rows of $\M{A}$.
Because the clustering partition splits the columns of $\M{A}$, each processor can partition its local data into left and right children to perform the split without any interprocessor communication.
If we use a 2D distribution for a given node, then because the partition is data dependent, a data redistribution is required in order to obtain a load balanced distribution of both children.
In the following subsections, we describe the parallel algorithms for Rank-2 NMF and approximating the principle singular value given this 1D data distribution and analyze their complexity in the context of the hierarchical clustering algorithm.

@@

@@

\begin{algorithm}
\caption{Parallel Power Method}
\label{alg:parpowmeth}
\begin{algorithmic}[1]
	\Require{$\M{A}$ is $m\times n$ and row-distributed across processors so that $\M[\hat]{A}$ is local $(m/p)\times n$ submatrix}
	\Function{$\sigma =$ Parallel-Power-Method}{$\M{A}$}
		\State Initialize $\V{v}$ randomly and redundantly
		\While{not converged}
			\State $\V[\hat]{u} = \M[\hat]{A} \V{v}$
			\State $\V[\hat]{z} = \M[\hat]{A}^\Tra \V[\hat]{u}$
			\State $\V{v} = \textsc{All-Reduce}(\V[\hat]{z})$
			\State $\sigma = \|\V{v}\|$
			\State $\V{v} = \V{v}/\sigma$
		\EndWhile
	\EndFunction
	\Ensure{$\sigma \approx \sigma_1(\M{A})$ is  redundantly owned by all procs}
\end{algorithmic}
\end{algorithm}

\begin{figure}
\input{fig/split}
\caption{Parallel splitting using Rank-2 NMF \GB{add more to caption and possibly color code W and H}}
\label{fig:split}
\end{figure}

>>>>

The communication cost of the algorithm depends on the parallel distribution of the input matrix data $\M{A}$.
In order to avoid redistribution of the matrix data, we choose a 1D row distribution so that each processor owns a subset of the rows of $\M{A}$.
Because the clustering partition splits the columns of $\M{A}$, each processor can partition its local data into left and right children to perform the split without any interprocessor communication.
If we use a 2D distribution for a given node, then because the partition is data dependent, a data redistribution is required in order to obtain a load balanced distribution of both children.
\Cref{fig:split} presents a visualization of the node-splitting process using a 1D processor distribution.
In the following subsections, we describe the parallel algorithms for Rank-2 NMF and approximating the principal singular value given this 1D data distribution and analyze their complexity in the context of the hierarchical clustering algorithm.

\begin{figure}
\input{fig/split}
\caption{Parallel splitting using Rank-2 NMF and 1D processor distribution.  A Rank-2 NMF computes factor matrices $\M{W}$ and $\M{H}$ to approximate $\M{A}$, the values of $\M{H}$ are used to determine child membership of each column (either red or blue), and the corresponding column of the $\M{W}$ matrix represents the part's feature weighting.  The 1D distribution is depicted for 3 processors to show that the splitting requires no interprocessor redistribution as the children are evenly distributed identically to the parent.}
\label{fig:split}
\end{figure}

@@

\Cref{alg:parrank2nmf} presents the parallelization of an alternating-updating scheme for NMF that uses the exact rank-2 solve algorithm presented in \Cref{alg:r2nnls} to update each factor matrix.
The algorithm computes the inputs to the rank-2 solves in parallel and then exploits the parallelism across rows of the factor matrix so that each processor solves for a subset of rows simultaneously.
The distribution of all matrices is 1D row distribution, so that each processor owns a subset of the rows of $\M{A}$, $\M{W}$, and $\M{H}$.
We use the notation $\M[\hat]{A}$ to refer to the $(m/p) \times n$ local data matrix and $\M[\hat]{W}$ and $\M[\hat]{H}$ to refer to the $(m/p) \times 2$ and $(n/p)\times 2$ local factor matrices.
With this distribution, the computation of $\M{W}^\Tra\M{W}$ and $\M{H}^\Tra\M{H}$ each is done via local multiplication followed by a single all-reduce collective.
All processors own the data they need to compute their contribution to $\M{A}^\Tra \M{W}$; in order to distribute the result to compute the rows $\M{H}$ independently, a reduce-scatter collective is used to sum and simultaneously distribute across processors.
To obtain the data needed to compute $\M[\hat]{W}$, each processor must access all of $\M{H}$, which is performed via an all-gather collective.
The iteration progresses until a convergence criterion is satisfied.
For performance benchmarking we use a fixed number of iterations, and in practice we use relative change in objective function value (residual norm).

@@

In order to compute the score for a frontier node, we use the difference between the principal singular value of the matrix columns of the node and the sum of those of its children.
Thus, we must determine the principal singular value of every node in the tree once, including leaf nodes.
We use the power method to approximate it, repeatedly applying $\M{A}\M{A}^\Tra$ to a vector until it converges to the leading right singular vector.
We present the power method in \Cref{alg:parpowmeth}.
Note that we do not normalize the approximate left singular vector so that the computed value approximates the square of the largest singular value.

Given the 1D distribution, only one communication collective is required for the pair of matrix-vector multiplications.
That is, the approximate right singular vector $\V{v}$ is redundantly owned on each processor, and the approximate left singular vector $\V{u}$ is distributed across processors.
Each processor can compute its local $\V[\hat]{u}$ from $\V{v}$ without communication and use the result for its contribution to $\V{v}=\M{A}^\Tra \V{u}$.
An all-reduce collective is used to obtain a copy of $\V{v}$ on every processor for the next iteration, and the norm is redundantly computed without further communication.
Again, we use a fixed number of iterations for benchmarking, but the more practical stopping criterion is based on the relative change in $\sigma$.

\begin{algorithm}
\caption{Parallel Power Method}
\label{alg:parpowmeth}
\begin{algorithmic}[1]
	\Require{$\M{A}$ is $m\times n$ and row-distributed across processors so that $\M[\hat]{A}$ is local $(m/p)\times n$ submatrix}
	\Function{$\sigma =$ Parallel-Power-Method}{$\M{A}$}
		\State Initialize $\V{v}$ randomly and redundantly
		\While{not converged}
			\State $\V[\hat]{u} = \M[\hat]{A} \V{v}$
			\State $\V[\hat]{z} = \M[\hat]{A}^\Tra \V[\hat]{u}$
			\State $\V{v} = \textsc{All-Reduce}(\V[\hat]{z})$
			\State $\sigma = \|\V{v}\|$
			\State $\V{v} = \V{v}/\sigma$
		\EndWhile
	\EndFunction
	\Ensure{$\sigma \approx \sigma_1^2(\M{A})$ is  redundantly owned by all procs}
\end{algorithmic}
\end{algorithm}
